Countering language attrition with PanLex and the Web of Data

### 2014-03-08 panlex-rdf-paper/2012/MLOD_PanLex/paper_v5.tex
Updated paper; still footnotes appear in references

The Semantic Web Initiative has led to the development of standards and technologies
supporting a machine-readable and \mbox{-i}nter\-pretable Linked Data
network, known as the \emph{Web of Data}.\footnote{\url{http://www.w3.org/standards/semanticweb/}} From these efforts the \emph{Linked Open Data (LOD) cloud}\footnote{\url{http://lod-cloud.net/}} emerged.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
A growing community is leveraging Semantic Web technologies for linguistic knowledge, building a \emph{Linguistic LOD (LLOD) cloud}.

Whenever there is Linked Data making use of descriptions, such as labels, in any language with lexemes in citation form (lemmas), the data are potentially linkable to, and through, PanLex. Here we describe how we connected PanLex to the Linked Data network.
In \autoref{sec:triplificate} we introduce the dataset, present a
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and, finally, in \autoref{sec:conclusion}, we summarize our recent and contemplated work.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
The PanLex \emph{database} is created by editors who consult \emph{information sources}\footnote{\url{http://panlex.org/tech/plrefs.shtml}}, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri. The data include single- and multi-word expressions, corresponding meanings assigned to them, and related information. While copies of the sources would not be interoperable, PanLex data constitute editors' \emph{interpretations} of sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema}.

@@

\begin{itemize}
  \item The \emph{source} entity is the authority to which an editor attributes assertions about lexical translations.
  \item \emph{Expressions} are lexical entities, each uniquely identified with a text, i.e. a string of (Unicode) characters, and a variety of a language. Expressions resemble lemmas or dictionary-entry headwords, but differ from them in at least two ways. (1) Homographs, such as the verb ``hide'' (conceal) and the noun ``hide'' (animal skin) in English, are treated as a single expression in PanLex. (2) Multiword expressions, such as ``fall in love'', traditionally found in an entry headed by one of their words, such as ``fall'' or ``love'', are treated as independent expressions in PanLex. Expression texts are constrained to comply with Unicode Normalization Form NFC.
    \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes, and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
    \item \emph {Language varieties} are collections of expressions. Each has a unique identifier: a language code and a distinguishing integer. For example, six dialects of Ahtna are identified as ``aht-000'' through ``aht-005''. These labels are, themselves, treated as a (controlled) language variety, whose expressions (i.e. the labels) are translated into natural languages and other controlled languages (such as the IETF standard BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}}).
%The only properties of this entity are ambiguity and synonymy: Controlled language varieties are usually defined to prohibit both of these.
    \item \emph{Meanings} are the entities assigned to expressions, thereby identifying expressions as translations or synonyms. For example, a source's translation of the German expression ``klingen'' into English as ``ring, sound, seem'' can be interpreted as (1) the assignment of a single meaning to all four expressions or (2) the assignment of two or three meanings to ``klingen'' and of one of those to each of the English expressions. Meanings are source-specific. The identification and consolidation of equivalent meanings of distinct sources is a research topic, not a database feature. Meanings can have properties of three types. (1) \emph{Definitions} are descriptions of a meaning, consisting of text strings annotated as being in particular language varieties. (2) \emph{Domains} are expressions (e.g., ``medicine'') that characterize a meaning, but do not express it. (3) \emph{Meaning identifiers} are strings acting as references to identifiers in a source.
    \item \emph{Denotations} are assignments of meanings to expressions. A denotation may have one or more \emph{word classes} (a closed set based on OLIF, the Open Lexicon Interchange Format) and/or \emph{metadata} (arbitrary strings paired as keys and values).
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} may define sources, attribute data to them, and define language varieties.
  \item Among the properties of sources are \emph{licenses}. Some of the license categories are \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (author invites inquiries), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (without a more liberal license), and \emph{PanLex Use Permission} (specific permission for use in PanLex).
    The distribution of licenses as of January 2014 is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations described above are the base for the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows.
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Source} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.46\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF vocabulary}
\label{fig:vocabulary}
\end{figure}

@@

\subsection{Transformation workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB), the recurrent conversion of the database to capture changes in it is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already reside in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view, or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach makes it easy to perform future revisions of RDF mapping, such as
adding support for new vocabularies.
%, as this does not require an extract transform load (ETL) process.
%and can therefore be done with little effort.
%This view-based approach makes it easy to revise the RDF mapping, such as adding support for new vocabular

@@

We intend to address some limitations in the future:
The relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.
The (assertion) source entity should be refactored to model users and information sources as distinct entities.
%Attribution is currently limited to (assertion) sources,
% obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are %themselves aggregations.
Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings,
but this sacrifices discovery possibilities when the metadata
%(as they often do)
describe facts that can again be expressed with PanLex expressions.
% (e.g., that the expression is archaic).
Finally, new collaborations between PanLex and related fields (e.g. as language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising arenas for development.

>>>>

The Semantic Web initiative has led to the development of standards and technologies
supporting a machine-readable and \mbox{-i}nter\-pretable Linked Data
network, known as the \emph{Web of Data}.\footnote{\url{http://www.w3.org/standards/semanticweb/}} From these efforts the \emph{Linked Open Data (LOD) cloud}\footnote{\url{http://lod-cloud.net/}} emerged.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
A growing community is leveraging Semantic Web technologies for linguistic knowledge, building a \emph{Linguistic LOD (LLOD) cloud}.

%Whenever there is Linked Data making use of descriptions, such as labels, in any language with lexemes in citation form (lemmas), the data are potentially linkable to, and through, PanLex.
Here we describe how we connected PanLex to this Linked Data network.
In \autoref{sec:triplificate} we introduce the dataset, present a
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} shows how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}, and related work is discussed in \autoref{sec:related}.
Finally, \autoref{sec:conclusion} concludes paper.
%In \autoref{sec:related} we discuss related work, and, finally, in \autoref{sec:conclusion}, we
% summarize our recent and contemplated work.
%conclude

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
The PanLex \emph{database} is created by editors who consult \emph{information sources}\footnote{\url{http://panlex.org/tech/plrefs.shtml}}, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri. The data include single- and multi-word expressions, corresponding meanings assigned to them, and related information.
%While copies of the sources would not be interoperable,
PanLex data constitute editors' \emph{interpretations} of sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema}.

@@

\begin{itemize}
  \item The \emph{source} entity is the authority to which an editor attributes assertions about lexical translations.
  \item \emph{Expressions} are lexical entities, each uniquely identified with a text, i.e. a string of (Unicode) characters, and a variety of a language. Expressions resemble lemmas or dictionary-entry headwords, but differ from them in at least two ways. (1) Homographs, such as the verb ``hide'' (conceal) and the noun ``hide'' (animal skin) in English, are treated as a single expression in PanLex. (2) Multiword expressions, such as ``fall in love'', traditionally found in an entry headed by one of their words, such as ``fall'' or ``love'', are treated as independent expressions in PanLex.
%Expression texts are constrained to comply with Unicode Normalization Form NFC.
% ~Claus: If we state this, we also have to cite NFC, which requires again more space
    \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes, and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
    \item \emph {Language varieties} are collections of expressions. Each has a unique identifier: a language code and a distinguishing integer. For example, six dialects of Ahtna are identified as ``aht-000'' through ``aht-005''. These labels are, themselves, treated as a (controlled) language variety, whose expressions (i.e. the labels) are translated into natural languages and other controlled languages (such as the IETF standard BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}}).
%The only properties of this entity are ambiguity and synonymy: Controlled language varieties are usually defined to prohibit both of these.
    \item \emph{Meanings} are entities assigned to expressions, thereby identifying expressions as translations or synonyms. For example, a source's translation of the German expression ``klingen'' into English as ``ring, sound, seem'' can be interpreted as (1) the assignment of a single meaning to all four expressions or (2) the assignment of two or three meanings to ``klingen'' and of one of those to each of the English expressions. Meanings are source-specific. The identification and consolidation of equivalent meanings of distinct sources is a research topic, not a database feature. Meanings can have properties of three types. (1) \emph{Definitions} are descriptions of a meaning, consisting of text strings annotated as being in particular language varieties. (2) \emph{Domains} are expressions (e.g., ``medicine'') that characterize a meaning, but do not express it. (3) \emph{Meaning identifiers} are strings acting as references to identifiers in a source.
    \item \emph{Denotations} are assignments of meanings to expressions. A denotation may have one or more \emph{word classes} (a closed set based on OLIF, the Open Lexicon Interchange Format) and/or \emph{metadata} (arbitrary strings paired as keys and values).
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} may define sources, attribute data to them, and define language varieties.
  \item Among the properties of sources are \emph{licenses}. Some of the license categories are \emph{public domain},
%\emph{Creative Commons (CC)},
\emph{request} (author invites inquiries),
%\emph{GNU General Public License (GPL)},
%\emph{GNU Lesser General Public License (LGPL)},
\emph{GNU Free Documentation License (FDL)},
%\emph{MIT License}, \emph{copyright} (without a more liberal license),
and \emph{PanLex Use Permission} (specific permission for use in PanLex).
    The distribution of licenses is shown in \autoref{tbl:plx-license-counts}.
%as of January 2014
\end{itemize}

@@

\subsection{The PanLex Vocabulary}
\label{sec:vocabulary}
The entities and relations described above are the base for the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small{http://ld.panlex.org/plx/}}, abbreviated with \texttt{\small{plx}}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows.
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Meanings and expressions are linked via \emph{denotations}.
%Following the semantics of the PanLex database,
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Source} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.46\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF Vocabulary}
\label{fig:vocabulary}
\end{figure}

@@

\subsection{RDF Transformation Workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB), the recurrent conversion of the database to capture changes in it is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already reside in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view, or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach makes it easy to perform future revisions of RDF mapping, such as
adding support for new vocabularies.
%, as this does not require an extract transform load (ETL) process.
%and can therefore be done with little effort.
%This view-based approach makes it easy to revise the RDF mapping, such as adding support for new vocabular

@@

We intend to address some limitations in the future:
The relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.
The source entity should be refactored to reference users and information sources as distinct entities.
%Attribution is currently limited to (assertion) sources,
% obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are %themselves aggregations.
Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings,
but this sacrifices discovery possibilities when the metadata
%(as they often do)
describe facts that can again be expressed with PanLex expressions.
% (e.g., that the expression is archaic).
Finally, new collaborations between PanLex and related fields (e.g. as language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising arenas for development.
